Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-22 paper/zhang-cook-laa-langrene-menendez.Rmd
Small changes in the first few sections.

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils information that would otherwise not be discovered and provides a more comprehensive understanding of the problem at hand. Task specific tools such as @li2020visualizing show how visualisation can be used to understand, for instance, the behaviour of neural network on classification models. However, no general visualisation tool is available for diagnosing optimisation procedures. The work presented in this paper brings visualization tools into optimisation problems with the aim to better understand the performance of optimisers in practice.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking, optimization can be unconstrained or constrained [@kelley1999iterative]. The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$\min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$\min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as multipliers and convex optimization methods including linear and quadratic programming can be used.

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis tool used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, characterised by the projection basis, that gives the most interesting low-dimensional view. A view is interesting if it can show some structures of the data, e.g., bi-modality, clustering and outliers, other than normal distribution.

The optimization challenges encountered in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple optima (local and global), the trade off between computational burden and proximity to the optima,  dealing with noisy objective functions that might be non-smooth and non-differentiable [@jones1998efficient]. Those  are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily applied to any other optimization problems.

The remainder of the paper is organised as follows. 
Section \ref{optim} provides an overview of optimisation methods, specifically line search methods. 
Section \ref{tour} reviews projection pursuit guided tour, defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data structure is defined to capture information during the optimisation, and different diagnostic plots are designed with the data collected.
Section \ref{application} shows applications of how these diagnostic plots can be used to discover interesting aspects of different optimisers and guide improvement to the existing algorithms.
Finally, Section \ref{implementation} describes the R package: \pkg{ferrn}, that implements the visual diagnostics.

@@

A large class of methods utilises the gradient information of the objective function, with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. As will be shown in the next section, the independent variable, in our optimisation problem, is a matrix of projection basis and the computational time required to perform differentation on a matrix could impede the render of tour animation. Hence, gradient-based methods are not the focus of this paper and considerations will be given to derivative-free methods.

Derivative-free methods, which do not rely on the knowledge of the gradient, are more generally applicable. Derivative-free methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Below will introduce three derivative-free methods, two of which are random search methods: creeping random search and simulated annealing, and the other one is pseudo derivative search.

Random search mehtods [@zabinsky2013stochastic] contain a random sampling component in its algorithm and has shown its ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sampling from the entire space would indicate a long time for convergence to occur despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have then been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating machanism. Creeping random search [@White1971-ci] is such a variation where a candidate point is generated within a neighbourhood of the current point. This gives creeping random search an advantage of being fast to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], takes a different updating machanism from the pure random search. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where nonimproving candidates can still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @belisle1992convergence; @granville1994simulated] and the global optimum can be attained under mild conditions.

Pseudo derivative search uses a common search scheme in optimistaion: line search. In line search methods, users are required to provide an initial estimate $x_{1}$ and, at each iteration, a search direction $S_k$ and a step size $\alpha_k$ are generated. Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desire convergence is reached. In derivative-free methods, local information of the objective function are used to determine the search direction. The choice of step size also needs considerations, as inadequate step sizes might prevent the optimisation method to converge to an optimum. An ideal step size can be chosen via finding the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ at each iteration.

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal and this has been argued for two reasons: Theoretical proofs from @diaconis1984asymptotics show that projections tend to be normal, as $n$ and $p$ approaching infinity under certain conditions. Index functions proposed in the litearture include early indices that can be categorised by measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indices [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

In projection pursuit the optimisation problem aims at finding the global and local maxima that give interesting projections according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

\noindent where $f$ and $\mathbf{X}$ are defined as in the previous section, $\mathcal{A}$ is the set of all $p$-dimensional projection bases, $I_d$ is the $d$-dimensional identity matrix, and the constraint ensures the projection bases, $\mathbf{A}$, to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the orthonormality constraint imposes a geometrical structure on the bases space: it forms a Stiefel manifold. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be computationally agile since the tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: Creeping random search (CRS), simulated annealing (SA), and pseudo derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of SA is summarised in Algorithm \ref{random-search} in the appendix. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SAJO gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even the landing basis does have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SAJO differs from SA in the inner loop.

Pseudo derivative (PD) search [@cook1995grand] uses a different strategy than SA and SAJO. Rather than randomly sample the basis space, PD first computes a search direction by evaluating vases close to the current basis. A step size is then chosen along the geodesic direction by another optimisation over an 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} in the appendix summarises the inner loop of the PD.

>>>>

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils insights that would otherwise not be discovered and provides a more comprehensive understanding of the problem at hand. Task specific tools such as @li2020visualizing show how visualisation can be used to understand, for instance, the behaviour of the optimisation for the example of neural network classification models. However, no general visualisation tool is available for diagnosing optimisation procedures. The work presented in this paper brings visualization tools into optimisation problems with the aim to better understand the performance of optimisers in practice.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is defined via an objective function. Broadly speaking, optimization can be unconstrained or constrained [@kelley1999iterative]. The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$\min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$\min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as multipliers and convex optimization methods including linear and quadratic programming can be used.

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis tool used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, represented by the projection matrix, that gives the most interesting low-dimensional view. A view is interesting if it can show some structures of the data other than a normal distribution, e.g., bi-modality, clustering and outliers.

The optimization challenges encountered in the projection pursuit guided tour problem are common to those of optimization in general. Examples include the existence of multiple optima (local and global), the trade-off between computational burden and proximity to the optima, or dealing with noisy objective functions that might be non-smooth and non-differentiable [@jones1998efficient]. The visualization tools and optimization methods presented in this paper can therefore easily be applied to other optimization problems.

The remainder of the paper is organised as follows. 
Section \ref{optim} provides an overview of optimisation methods, specifically line search methods. 
Section \ref{tour} reviews projection pursuit guided tour, defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data structure is defined to capture information during the optimisation, and different diagnostic plots are designed with the data collected.
Section \ref{application} shows applications of how these diagnostic plots can be used to discover interesting aspects of different optimisers and guide improvements to the existing algorithms.
Finally, Section \ref{implementation} describes the R package: \pkg{ferrn}, that implements the visual diagnostics.

@@

A large class of methods uses the gradient information of the objective function, with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the limited complexity of the constraints. As will be shown in the next section, the independent variables in our optimisation problem are the entries of a projection matrix and the computational time required to perform differentiation on a matrix could impede the render of tour animation. Hence, gradient-based methods are not the focus of this paper and considerations will be given to derivative-free methods.

Derivative-free methods, which do not rely on the knowledge of the gradient, are more generally applicable. Derivative-free methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Here we consider three derivative-free methods, two of which are random search methods: creeping random search and simulated annealing, and the other one is pseudo derivative search.

Random search methods [@zabinsky2013stochastic] have a random sampling component as part of their algorithm and have been shown to have the ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sufficient sampling from the entire space would require a long time for convergence to occur, despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have thus been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating machanism. Creeping random search [@White1971-ci] is such a variation, where a candidate point is generated within a neighbourhood of the current point. This makes creeping random search faster to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], introduces a different updating mechanism. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where worse candidates still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @belisle1992convergence; @granville1994simulated] and the global optimum can be attained under mild conditions.

Pseudo derivative search uses a common search scheme in optimistaion: line search. In line search methods, users are required to provide an initial estimate $x_{1}$ and, at each iteration, a search direction $S_k$ and a step size $\alpha_k$ are generated. Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desired convergence is reached. In derivative-free methods, local information of the objective function is used to determine the search direction. The choice of step size also needs consideration, as inadequate step sizes might prevent the optimisation method to converge to an optimum. An ideal step size can be chosen by finding the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ at each iteration.

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal and this has been argued for two reasons: Theoretical proofs from @diaconis1984asymptotics show that projections tend to be normal, as $n$ and $p$ approach infinity under certain conditions. Index functions proposed in the litearture include early indices that can be categorised by measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indices [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different versions depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

In projection pursuit the optimisation aims at finding the global and local maxima that give interesting projections according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

\noindent where $f$ and $\mathbf{X}$ are defined as in the previous section, $\mathcal{A}$ is the set of all $p$-dimensional projection bases, $I_d$ is the $d$-dimensional identity matrix, and the constraint ensures the projection bases, $\mathbf{A}$, to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the orthonormality constraint imposes a geometrical structure on the bases space: it forms a Stiefel manifold. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be fast to compute since the tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: Creeping random search (CRS), simulated annealing (SA), and pseudo derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of CRS is summarised in Algorithm \ref{random-search} in the appendix. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SA gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even if the candidate basis does not have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SA differs from CRS in the inner loop.

Pseudo derivative (PD) search [@cook1995grand] uses a different strategy than CRS and SA. Rather than randomly sample the basis space, PD first computes a search direction by evaluating bases close to the current basis. A step size is then chosen along the corresponding geodesic by another optimisation over an 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} in the appendix summarises the inner loop of the PD.
